---
title: "BF_Midterm"
author: "Ritik Kumar"
date: "2024-11-04"
output: html_document
---

## Importing the Data
```{r}
library(fpp)
library(fpp2)
library(ggplot2)
library(forecast)
library(readr)
library(dplyr)

sales_data <- read_csv("sales.csv")
head(sales_data)
```

## Introduction about Dataset 
* The dataset represents the sales of a healthcare product manufactured by a pharmaceutical company in the U.S. Because the production process for this drug is lengthy and time-consuming, it's crucial to forecast sales accurately. By predicting future sales, we can optimize the manufacturing schedule, ensuring that production aligns closely with demand. This forecasting approach helps the company maintain a steady supply of the drug, allowing customer orders to be fulfilled immediately.
* First we are successfully importing the data.
* Then we are displaying the first few rows for confirmation that are dataset is imported successfully

## Plot and Inference
##### Converting data into Times Series and then Plotting it
```{r}
sales_ts <- ts(sales_data$Sales, frequency = 12, start = c(2020,1), end=c(2024,8))
plot(sales_ts, main="Sales of a healthcare product manufactured by a US pharma company", xlab="Year", ylab="Sales")
```

* we start with plotting the time series for visualization and understand the data correctly

##### Initial Oberversations of time series plot
* The plot shows that there is an increasing trend in the sales starting from year Jan 2020 till around 2023 August.
* From August 2023 till August 2024, there has been a decreasing trend in the sales.
* From 2023, there has been a seen a increasing and decreasing fluctuation trend till the year 2024
* The data however appear to show some strong seasonal variation starting from 2023. 
* from 2020 it will be more than 5 years data that we are considering and this data should be good enough to be considered for forcasting.

## Central Tendency

##### Min, max, mean, median, 1st and 3rd Quartile values of the times series
```{r}
summary(sales_ts)
```

* The summary function above gives the min, max, mean, median, 1st and 3rd Quartile values of the times series.

##### Show the Box Plot
```{r}
boxplot(sales_ts, main ='Boxplot for the Sales Time Series')
Acf(sales_ts)
```

##### Observations and Inferences 
* The boxplot shows that there are no outliers in the data.
* The median is more towards the first quartile.
* From summary, we can also see that the median value is less then the mean value for the time series.
*  This means that the data is right skewed. This can be justified seeing the boxplot as well.
* From the ACF plot, we can see that many of the values crossed the confidence intervals, stating there is a trend component in the data.
* Also, we can see that after 18th lag period, the ACF plot is dipping into the negative values stating seasonality also exists in the data.

## Decomposition

##### Decomposition Plot
```{r}
stl_decomp <- stl(sales_ts, s.window = "periodic")
plot(stl_decomp, main = 'Decomposition Plot')
```

##### Is the time series seasonal?
* Yes, the time series is seasonal.

##### Is the decomposition additive or multiplicative?
```{r}
decom <- decompose(sales_ts)
decom$type
```

* The decomposition seems to be additive. 
* Because, with as trend increases, we do not see any increase in the seasonality. The seasonality appears to be the same throughout.

##### If seasonal, what are the values of the seasonal monthly indices? 
```{r}
decom$figure
```

##### Observations and Inferences

* From 2020 to 2024, the values of the time series seem to increase throughout.
* We can see a peak in March 2020 and then a dip and June 2020 and then continuous decrease.
* Then again a similar case for Jan 2024 and then a dip in March 2024 and then a continuous decrease.

##### Seasonality adjusted plot
```{r}
plot(sales_ts, ylab = 'Sales')
lines(seasadj(stl_decomp), col="Red")
```

* There are high fluctuations that can be observed after applying seasonal adjustment.
* With time, these fluctuations will cause more deviations and change our forecast. So, it is important to consider the seasonal variation in the data.

## Testing the various Forecasting methods for the given dataset

## Naive Method 

##### Output
```{r}
naive_forecast <- naive(sales_ts)
plot(naive_forecast)
```

##### Perform Residual Analysis for this technique. 
```{r}
plot(naive_forecast$residuals)
```

* The residuals appear to have increasing positive values and then peaked in the third quarter of the year 2023 and then dipped down.
* All the residuals are positive. The residuals seems to be little close to mean at zero.

###### Residuals Histogram
```{r}
hist(naive_forecast$residuals)
```

* The histogram appears to be normally distributed.
* But the values do not have a mean zero. Although,the histogram appears to be normally on both side.
* This means that the data is biased as the mean is not zero.


###### Fitted vs Residual Values
```{r}
cbind(Fitted = fitted(naive_forecast), 
      Residuals = residuals(naive_forecast)) %>%
  as.data.frame() %>%
  ggplot( aes(x= Fitted, y= Residuals)) + geom_point()
```

* The Fitted vs Residuals plot appears to have a trend.
* Also, all the points are starting from mean 0.
* This means there is heteroscedasticity in the errors which means that the variance of the residuals may not be constant.

###### Actual vs Residual values
```{r}
cbind(Actual=sales_ts,
      Residuals=residuals(naive_forecast)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()
```

* Similar to the previous plot, The actual vs Residuals plot also appears not to be random. 

###### ACF of residuals
```{r}
Acf(naive_forecast$residuals)
```

* Some values of the Acf have crossed the confidence level meaning there is a trend in the residuals and we have missed some variable in our forecast.
* The Acf values also show seasonality in the plot and we missed this variable too.
* Meaning that naive forecast is missing some main variables which we have missed our consideration for the forecast.

##### Accuracy
```{r}
accuracy(naive_forecast)
```

##### Forecast 
```{r}
forecast(naive_forecast)
plot(forecast(naive_forecast))
```

##### Naive Method Summary
* The ME, RMSE values are very high indicating that this method may not be the right one to go with.
* We can consider more forecasting techniques and check if the error values are less than this one.
* From Jan 2020 to Jan 2024 there is observed to be an increasing trend in the data. So, naive forecast may not be a right way to forecast.
* Also, the predicted forecast also seems to be stable.

## Simple Moving Averages

##### Plot the graph for time series. 
##### Simple Moving average of order 3, 6, and 9
```{r}
ma3_forecast = ma(sales_ts, order = 3)
ma6_forecast = ma(sales_ts, order = 6)
ma9_forecast = ma(sales_ts, order = 9)
plot(sales_ts, main ="Plot along with moving averages")
lines(ma3_forecast, col='Red')
lines(ma6_forecast, col='Blue')
lines(ma9_forecast, col='Green')
```

##### Show the forecast for the next 12 months using one of the simple averageorders that you feel works best for time series
```{r}
ma3_forecast = ma(sales_ts, order = 3)
plot(sales_ts, main ="Plot along with moving averages for 12 Months", h=12)
lines(ma3_forecast, col='Red')
```

##### Observations

* From the plots, it is observed that the higher the order we consider, the smoother the moving average curve in the plot.
* It can be seen that the Green line above is the smoothest compared to Blue or Red lines.
* The Red line (order 3) gives the most real data compared to the other two. The higher order averages smoother the plot and do not give the actual values.

## Simple Smoothing
```{r}
ses_data = ses(sales_ts)
plot(ses_data)
attributes(ses_data)
```

```{r}
summary(ses_data)
```

##### Observations 
* alpha = 0.4319
* Alpha specifies the coefficient for the level smoothing. Values near 1.0 mean that the latest value has more weight.
* Initial state: l = 1472.029
* Sigma: 2603.101 Sigma defines the variance in the forecast predicted.

##### Residual Analysis
```{r}
plot(ses_data$residuals)
```

* The residuals appear to have increasing positive values and then peaked in the third quarter of the year 2023 and then dipped down.
* Most of the residual values appear to be positive and do not have a mean of zero.

###### Histogram plot of residuals
```{r}
hist(ses_data$residuals)
```

* The histogram appears to be normally distributed.
* But the values do not have a mean zero. The histogram appears to be both side normally.

###### Fitted values vs. residuals
```{r}
cbind(Fitted=fitted(ses_data),
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted,y=Residuals)) + geom_point()
```

* The Fitted vs Residuals plot appears to have a trend. 
* This means there is heteroscedasticity in the errors which means that the variance of the residuals may not be constant.

###### Actual values vs. residuals
```{r}
cbind(Actual = sales_ts,
      Residuals=residuals(ses_data)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()
```

* Similar to the previous plot, the Actual vs. Residuals plot appears to have some trend in the data.

###### ACF plot of the residuals
```{r}
Acf(ses_data$residuals)
```

* Some Values of the Acf have crossed the confidence level meaning there is a trend in the residuals and we have missed some variable in our forecast.
* The Acf values also show seasonality in the plot and we missed this variable too.
* Meaning that simple smoothing is missing some main variables which we have missed our consideration for the forecast.

##### Accuracy
```{r}
accuracy(ses_data)
```

##### Forecast 
```{r}
ses_data
plot(ses_data)
```

##### Simple Smoothing Summary

* The ME, RMSE values are very high indicating that this method may not be the right one to go with.
* We can consider more forecasting techniques and check if the error values are less than this one.
* From Jan 2020 to Jan 2024 there is observed to be an increasing trend in the data. So, this forecast may not be a right way to forecast.
* We can try Holtwinters approach as it suits for trend+seasonal time series

##  Holt-Winters 
```{r}
Hw_forecast <- hw(sales_ts, seasonal = "additive")
plot(forecast(Hw_forecast))
attributes(Hw_forecast)
Hw_add <- forecast(Hw_forecast)
```

* Here, additive Holtwinters method is considered.
* This is because the seasonality isn't increasing with trend. This is an additive time series.

###### Observations
```{r}
Hw_add$model
```

* Alpha = 0.3176 Alpha specifies the coefficient for the level smoothing in Holtwinters.
* Beta = 0.0862 Beta specifies the coefficient for the trend smoothing in Holtwinters. 
* Gamma = 1e-04  Gamma specifies the coefficient for the seasonal smoothing in Holtwinters.
* Values 1.0 means that the latest value has highest weight.
* Initial states:
    l = 818.4249 
    b = 343.2258 
    s = 1774.076 -1565.1 -1259.728 1059.152 -814.7662 753.5157
           2154.785 120.2898 -57.6465 1068.374 -1562.394 -1670.559
* Sigma = 2428.542 Sigma defines the variance of the forecast values.

##### Residual Analysis
```{r}
plot(Hw_add$residuals)
```

* The residuals appear to be random and also the mean looks to be near zero. We can check this with histogram.
* We can observe a couple of up and downs throughout. But even they did not show and growing residual pattern.

###### Histogram plot of residuals
```{r}
hist(Hw_add$residuals)
```

* The histogram appears to be normally distributed.
* And the mean is near zero. Indicating the data is not biased.
* Overall, comparing the previous forecasts, this forecast appears to be the best till now. 


###### Fitted values vs. residuals
```{r}
cbind(Fitted = fitted(Hw_add),
      Residuals=residuals(Hw_add)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Fitted, y=Residuals)) + geom_point()
```

* The Fitted vs Residuals plot appears not to have any trend.
* This means there is no heteroscedasticity in the errors which means that the variance of the residuals is constant.


###### Actual values vs. residuals
```{r}
cbind(Actual= sales_ts,
      Residuals=residuals(Hw_add)) %>%
  as.data.frame() %>%
  ggplot(aes(x=Actual, y=Residuals)) + geom_point()
```

* Similar to the previous plot, the Actual vs. Residuals plot appears to be random.

###### ACF plot of the residuals
```{r}
Acf(Hw_add$residuals)
```

* In the Acf plot, none of the values crossed the confidence levels. It appears to be white noice.
* This signifies that the forecast is a good forecast.
* This proves to be the best forecast comparing all the previous ones tested.

##### Accuracy
```{r}
accuracy(Hw_add)
```

##### Forecast 
```{r}
forecast(Hw_forecast)
plot(forecast(Hw_forecast))
```

##### Holtwinters Summary
* The ME, RMSE values are quite low compared to any of our previous forecasts.
* Holwinters is a better forecast compared to naive and simple smoothing.
* Holtwinters appears to be the best forecast considering all the previous forecast methods.

## Accuracy Summary
```{r}
accuracy(naive_forecast)
accuracy(ses_data)
accuracy(Hw_add)
```

##### Best & Worst Forecasts

* To start with, there is nothing like best or worst forecast.
* Considering the accuracy data above, HoltWinters forecast seems to fit the time series the best as it has the least error values.
* And naive forecast seems to be the worst as it has the largest ME and RMSE values.

## Conclusion

* The data seemed to have trend and seasonality initially and we checked the same with Acf and confirmed it.
* Based on the three forecasting methods naive, simple smoothing, and HoltWinters, we can see that HoltWinters forecast provides to be the better forecasting method in this case.
* This is because the forecast fits perfectly and also the error values are quiet low for HoltWinters forecast.
* Additionally residuals in HoltWinters appear to be random and the all the ACF values of residuals are within the confidence interval.
* This shows that our hypothesis is correct based on the accuracy of all the models.
* Based on the analysis and forecast, the time series will decrease over the next year and the next 2 years.